Modelling the emergence of whisker barrels

### 2020-01-24 paper/briefpaper/brief.tex
Re-wraps the bioRXiv version of the paper, then applies eLife formatting and some initial changes

% It's about how a somatotopically ordered pattern can be
% re-established in the cortex based on information carried with
% thalamocortical axons afferent from the ventrobasal thalamus, but
% without specifying absolute positions as center-points for the
% barrels. Everything in the paper should be tailored towards arguing
% this idea.

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
\templatetype{pnasresearcharticle}

\title{Modelling the emergence of whisker barrels}
% While writing, don't stop for errors
%\nonstopmode

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Sebastian~S.~James}
\author[b]{Leah~A.~Krubitzer}
\author[a]{Stuart~P.~Wilson}

\affil[a]{Department of Psychology, The University of Sheffield, Sheffield, United Kingdom.}
\affil[b]{Center for Neuroscience, The University of California, Davis, United States.}
% $^{*}$Corresponding author email: Seb.James@Sheffield.ac.uk\vspace{1em}\\

\leadauthor{James}

\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: \href{mailto:seb.james@sheffield.ac.uk}{\nolinkurl{seb.james@sheffield.ac.uk}}}

\keywords{Barrel cortex $|$ Self-organization $|$ Somatotopic map $|$ Axon guidance}

@@

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

% Blue for comments
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\e}{\emph}
\newcommand{\bol}{\textbf}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\makeatother

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\modulolinenumbers{}
\linenumbers

\dropcap{S}patial patterns in neural connectivity provide clues about the
constraints under which brains evolve and develop
\citep{purves_iterated_1992}. A distinctive example is found in the barrel
cortex of many rodent species. The barrels become identifiable in layer 4 of
primary somatosensory cortex as clusters of thalamocortical axons, enclosed by
borders a few neurons thick from postnatal day 3
\citep{erzurumlu_development_2012}. Across the cortical surface the barrels
constitute a somatotopic map of the whiskers, with cells within adjacent
barrels responding most strongly and quickly to deflection of adjacent
whiskers \citep{armstrong-james_flow_1992}. Barrel patterning reflects
subcortical whisker maps comprising cell aggregates called barrelettes in the
brainstem and barreloids in the thalamus.
%\cite{van_der_loos_barreloids_1976}.

Barrel formation requires afferent input from whisker stimulation and thalamic
calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex
network of axon guidance molecules such as ephrin-A5 and A7 and adhesion
molecules such as cadherin-6 and 8 \citep{miller_epha7-ephrin-a5_2006}.  This
network is orchestrated by interactions between morphogens Fgf8 and Fgf17 and
transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005}, which are expressed in gradients that mark
orthogonal axes and can be manipulated to stretch, shrink, shift, and even
duplicate barrels \cite{assimacopoulos_fibroblast_2012}.
% Removed ref: ,bishop_regulation_2000 after ..Coup-tf1

The barrel boundaries form a Voronoi tessellation across the cortical sheet
\citep{senft_mouse_1991} (Fig.\,1A), suggesting that barreloid topology is
preserved in the projection of thalamocortical axons into the cortex, and that
a barrel forms by lateral axon branching from an initial center-point that
ceases upon contact with axons branching from adjacent centers. But the
assumption of pre-arranged centers is difficult to resolve with the
observation that axons arrive in the cortical plate as an undifferentiated
bundle, prior to barreloid formation \cite{agmon_organized_1993}.

Alternatively, reaction-diffusion dynamics could generate a Voronoi
tessellation without pre-arranged centers, by amplifying characteristic modes
in a noisy initial distribution of axon branches, as a net effect of
short-range cooperative and longer-range competitive
interactions. Accordingly, the barrel pattern would be determined by the
relative strength of these interactions and by the shape of the cortical field
boundary. However, intrinsic cortical dynamics alone cannot account for the
topographic correspondence between thalamic and cortical domains, the
irregular sizes and specific arrangement of the barrels, or the influence of
gene expression gradients.

The center-point and reaction-diffusion models are not mutually
exclusive. Pre-organised centers could bias reaction-diffusion processes to
generate specific arrangements more reliably, and mechanisms of lateral axon
branching may constitute the tension between cooperation and competition
required for self-organization. However, proof that barrel patterning can
emerge from an undifferentiated bundle of axons, based only on local
interactions, would show that a separate mechanism for pre-organising
thalamocortical connections need not be assumed. To this end, we ask whether
barrel maps can emerge in a system with reaction-diffusion dynamics, under the
guidance of signalling gradients, and in the absence of pre-defined centers.

\section*{Models}

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a
reaction-diffusion style model of how extrinsic signalling gradients can
constrain the emergence of distinct fields from intrinsic cortical
dynamics. Their model defines how the density of connections $c(x,t)$ and axon
branches $a(x,t)$ interact at time $t$, along a 1D anterior-posterior axis
$x$, for $N$ thalamocortical projections indexed by $i$. The model was derived
from the assumption that the rates at which $a_i$ and $c_i$ grow are
reciprocally coupled. Extending the original 1D model to simulate arealization
on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and model
synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. The second term on the right represents
an assumption that axon branching decreases when synaptogenesis increases, and
vice versa, i.e., that axon branches are required for synaptic connections to
form. Here $\chi_i=0$ is a placeholder.

%All software required to reproduce these results is available at \url{https://github.com/ABRG-Models/BarrelEmerge}.

\section*{Results}

\begin{figure*}
\begin{center}
\includegraphics[width=17.8cm]{./MainFig.png}
\end{center}
\caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
  by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
  measure departure (Honda-$\delta$; see \cite{senft_mouse_1991}) from a
  perfect Voronoi tesselation. Right shows the initial distribution of axon
  branching density ($a$) for one thalamocortical projection, and two
  molecular guidance fields ($\rho$). \textbf{B} The strengths of interaction
  $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
  projections by the lengths of green and blue arrows respectively, assuming
  that similar fields aligned to the posterior-anterior and medial-lateral
  axes in the ventroposterior medial nucleus of the thalamus are sampled at
  the locations of putative barreloid centers (reconstructed from
  \cite{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
  $N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
  $\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
  projection for which the connection density is maximal, black lines
  delineate boundaries, and overlaid contours show $c>0.5$ (see Movie
  S1). Software for modelling the cortical sheet as a hexagonal lattice with
  no-flux boundary conditions and an arbitrary boundary shape, for computing
  the flux using a finite difference method, for numerical integration by the
  fourth-order Runge-Kutta method, and for reproducing these results is
  available at \url{https://github.com/ABRG-Models/BarrelEmerge} \textbf{D} Red
  dots show the Honda-$\delta$ metric obtained from simulation approaching
  that obtained from barrels in \textbf{A} (dotted line); black squares
  measure the correspondence between the real and simulated barrel shapes,
  $\eta$ (the product of the sum of squared differences between real and
  simulated centers and the sum of differences in area; units
  mm$^4$). \textbf{E} Guidance fields and emergent barrel pattern in a Fgf8
  misexpression experiment (c.f. \cite{assimacopoulos_fibroblast_2012}),
  simulated by reflecting $\rho_1$ at the join ($\epsilon=80$). All scale bars
  1mm.}
\label{fig:main}
\end{figure*}

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced in our 2D implementation. Using an
elliptical boundary with $M=3$ offset guidance gradients aligned to the longer
axis, $N=5$ projections gave rise to five distinct cortical fields at
locations that preserved the topographic ordering defined by the original
$\gamma$ values. However, we found that specifying $N$ ordered areas required
$M\approx (N+1)/2$ signalling fields. As the number of distinct guidance
fields is unlikely to approach the number of barrels, further extensions were
required.

The term in parentheses in Eq.~\ref{eq:dc} represents competition between
projections for limited cortical connections. To introduce axon branching
competition we redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial density at $t=0$;
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma},
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$.

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that a projection interacts with a given ephrin field in
the cortical subplate with a strength determined by the concentration of a
similar molecule at its thalamic origin, i.e., the putative barreloid
center. As such, two orthogonal linear thalamic gradients were defined, from
which 41 pairs of $\gamma$ values were sampled, at the coordinates of 41
barreloid centers recreated from \cite{haidarliu_size_2001} (Fig.\,1B).
%
A cortical boundary enclosing 41 corresponding barrels was traced from a
cytochrome oxidase stain from \cite{zheng_signal_2001}, and
Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes. From random initial conditions for
$a({\bf x},0)$, a clear Voronoi-like tesselation emerged (Fig.\,1C; Movie
S1). The Honda $\delta$ metric (see \cite{senft_mouse_1991}) confirmed that a
`good' Voronoi pattern emerged within $\approx 3000$ iterations (Fig.\,1D). A
measure of the difference in shape between real and simulated barrels revealed
a strong correspondence (see Fig.\,1D).

To further investigate the interplay of genetic and intrinsic mechanisms we
simulated a seminal barrel duplication experiment
\cite{shimogori_fibroblast_2005}, by solving on a domain comprising two
separate, identically shaped cortical boundaries. In this case one distinct
cortical field per thalamic projection emerged, within one of the two
boundaries only, i.e., no duplication. However, merging the two fields to
create an extended mirror-symmetric boundary shape, with an anterior-posterior
guidance field reflected at the join, gave rise to two mirror-symmetrical
barrel fields comprising $\approx 2N$ barrels, i.e., two cortical fields for
each projection (Fig.\,1E). Together these results suggest that by the
misexpression of Fgf8, Shimogori \& Grove \citep{shimogori_fibroblast_2005}
effectively created one large, rather than two distinct, cortical fields.

% Omit to shorten manuscript
%\subsection*{Movie S1 caption}
%
%Movie corresponding to Fig.\,1C in the main paper. Simulation parameters were
%$N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
%$\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
%projection for which the connection density is maximal, black lines delineate
%boundaries, and overlaid contours show $c>0.5$. The final frame in the movie
%is step 25000 of the simulation.

\section*{Discussion}

These results suggest that the emergence of realistic barrel patterning
requires that i) at each cortical location thalamocortical projections compete
for a limited number of available synaptic connections
(Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a
given projection is reduced by the density of other projections
(Eq.~\ref{eq:comp}), and iii) the branch density of each projection is
conserved over time (Eq.~\ref{eq:norm}).

The emergence of barrels in simulation required competition between
projections for synaptic connectivity, and also for cortical space
(represented by $\chi$). This requirement may account for the absence of
barrels in rodents with larger brains, such as capybara, for which competition
for space is presumably weaker \citep{Woolsey1975}. Hence, irrespective of
whether barrels are necessary for adaptive whisker function, the emergence of
somatotopically ordered modular structures may be an inevitable consequence of
local competition for cortical territory driven by input from an array of
discrete sensory organs \citep{purves_iterated_1992}.

We emphasize that the formulation of the model is entirely local; the
simulation permits information to be communicated from a given cortical grid
cell only to those immediately adjacent (via diffusion). Hence the simulations
demonstrate how a self-organising system, constrained by genetically specified
guidance cues and by the shape of the boundary, can faithfully reproduce an
arrangement of cell aggregates in one neural structure as a topographic map in
another.

Moreover, the present results confirm that somatotopic map formation does not
require the pre-specification of center-points by as yet undetermined
additional developmental mechanisms.

\acknow{The authors thank Jason Berwick for advice and data used to construct
  Fig.\,1A. This work was supported by the James S.~McDonnell Foundation
  (grant 220020516).}

\showacknow{} % Display the acknowledgments section

>>>>

% It's about how a somatotopically ordered pattern can be
% re-established in the cortex based on information carried with
% thalamocortical axons afferent from the ventrobasal thalamus, but
% without specifying absolute positions as centre-points for the
% barrels. Everything in the paper should be tailored towards arguing
% this idea.

\documentclass[9pt,lineno]{elife}

\usepackage{amsmath,esint}
\newcommand*\dif{\mathop{}\!\mathrm{d}}

\title{Modelling the emergence of whisker barrels}

\author[1*]{Sebastian~S.~James}
\author[2]{Leah~A.~Krubitzer}
\author[1]{Stuart~P.~Wilson}

\affil[1]{Department of Psychology, The University of Sheffield, Sheffield, United Kingdom.}
\affil[2]{Center for Neuroscience, The University of California, Davis, United States.}
\corr{seb.james@sheffield.ac.uk}{SSJ}

%\keywords{Barrel cortex $|$ Self-organization $|$ Somatotopic map $|$ Axon guidance}

\maketitle

@@

% Blue for comments
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\e}{\emph}
\newcommand{\bol}{\textbf}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}

\section{Introduction}

Spatial patterns in neural connectivity provide clues about the constraints
under which brains evolve and develop \citep{purves_iterated_1992}. Perhaps
the most distinctive pattern can be found in the barrel cortex of many rodent
species \cite{woolsey_structural_1970}. The barrels are identifiable soon
after birth in layer 4 of primary somatosensory cortex as dense clusters of
thalamocortical axons, which are enclosed by borders a few neurons thick from
postnatal day 3 \citep{erzurumlu_development_2012}. In the plane tangential to
the cortical surface the barrels constitute a somatotopic map of the whiskers,
with cells within adjacent barrels responding most strongly and quickly to
deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. Barrel
patterning reflects subcortical whisker maps comprising cell aggregates called
barrelettes in the brainstem and barreloids in the thalamus.
%\cite{van_der_loos_barreloids_1976}.

Barrel formation requires afferent input from whisker stimulation and thalamic
calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex
network of axon guidance molecules such as ephrin-A5 and A7 and adhesion
molecules such as cadherin-6 and 8
\citep{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}.  This network
is orchestrated by interactions between morphogens Fgf8 and Fgf17 and
transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005}, which are expressed in gradients that mark
orthogonal axes and can be manipulated to stretch, shrink, shift, and even
duplicate barrels \cite{assimacopoulos_fibroblast_2012}.
% Removed ref: ,bishop_regulation_2000 after ..Coup-tf1

The barrel boundaries form a Voronoi tessellation across the cortical sheet
\citep{senft_mouse_1991} (Fig.\,1A), suggesting that barreloid topology is
preserved in the projection of thalamocortical axons into the cortex, and that
a barrel forms by lateral axon branching from an initial centre-point that
ceases upon contact with axons branching from adjacent centres.  \cmnt{Another
  sentence about center-points here.} But the assumption of pre-arranged
centres is difficult to resolve with the observation that axons arrive in the
cortical plate as an undifferentiated bundle, prior to barreloid formation
\cite{agmon_organized_1993}.

Alternatively, reaction-diffusion dynamics could generate a Voronoi
tessellation without pre-arranged centres, by amplifying characteristic modes
in a noisy initial distribution of axon branches, as a net effect of
short-range cooperative and longer-range competitive
interactions. Accordingly, the barrel pattern would be determined by the
relative strength of these interactions and by the shape of the cortical field
boundary. However, intrinsic cortical dynamics alone cannot account for the
topographic correspondence between thalamic and cortical domains, the
irregular sizes and specific arrangement of the barrels in rows and arcs, or
the influence of gene expression gradients.

The centre-point and reaction-diffusion models are not mutually
exclusive. Pre-organised centres could bias reaction-diffusion processes to
generate specific arrangements more reliably, and mechanisms of lateral axon
branching may constitute the tension between cooperation and competition
required for self-organization. However, proof that barrel patterning can
emerge from an undifferentiated bundle of axons, based only on local
interactions, would show that a separate stage and/or extrinsic mechanism for
pre-organising thalamocortical connections need not be assumed. To this end,
we ask whether barrel maps can emerge in a system with reaction-diffusion
dynamics, under the guidance of signalling gradients, and in the absence of
pre-defined centres.

\section{Models}

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a
reaction-diffusion style model of how extrinsic signalling gradients can
constrain the emergence of distinct fields from intrinsic cortical
dynamics. Their model defines how the density of connections $c(x,t)$ and axon
branches $a(x,t)$ interact at time $t$, along a 1D anterior-posterior axis
$x$, for $N$ thalamocortical projections indexed by $i$. The model was derived
from the assumption that the rates at which $a_i$ and $c_i$ grow are
reciprocally coupled. Extending the original 1D model to simulate arealization
on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and model
synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. The second term on the right represents
an assumption that axon branching decreases when synaptogenesis increases, and
vice versa, i.e., \cmnt{when a connection is made, the axon branching
  decreases by one branch and when a connection is lost, one branch is added
  to the branching density}
%that axon branches are required for synaptic connections to form.
Here $\chi_i=0$ is a placeholder.

Software for modelling the cortical sheet as a hexagonal lattice with no-flux
boundary conditions and an arbitrary boundary shape, for computing the flux
using a finite difference method, for numerical integration by the
fourth-order Runge-Kutta method, and for reproducing these results is
available at \url{https://github.com/ABRG-Models/BarrelEmerge}.

\section{Results}

\begin{figure}
\begin{fullwidth}
\includegraphics[width=\linewidth]{./MainFig_superlowres.png}
\caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
  by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
  measure departure (Honda-$\delta$; see \cite{senft_mouse_1991}) from a
  perfect Voronoi tesselation. Right shows the initial distribution of axon
  branching density ($a$) for one thalamocortical projection, and two
  molecular guidance fields ($\rho$). \textbf{B} The strengths of interaction
  $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
  projections by the lengths of green and blue arrows respectively, assuming
  that similar fields aligned to the posterior-anterior and medial-lateral
  axes in the ventroposterior medial nucleus of the thalamus are sampled at
  the locations of putative barreloid centers (reconstructed from
  \cite{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
  $N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
  $\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
  projection for which the connection density is maximal, black lines
  delineate boundaries, and overlaid contours show $c>0.5$ (see Movie
  S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from
  simulation approaching that obtained from barrels in \textbf{A} (dotted
  line); black squares measure the correspondence between the real and
  simulated barrel shapes, $\eta$ (the product of the sum of squared
  differences between real and simulated centers and the sum of differences in
  area; units mm$^4$). \textbf{E} Guidance fields and emergent barrel pattern
  in a Fgf8 misexpression experiment
  (c.f. \cite{assimacopoulos_fibroblast_2012}), simulated by reflecting
  $\rho_1$ at the join ($\epsilon=80$). All scale bars 1mm.}
\label{fig:main}
\end{fullwidth}
\end{figure}

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced using our extension to a 2D cortical
sheet. Using an elliptical domain, $S$, with $M=3$ offset guidance gradients
aligned to the longer axis, $N=5$ thalamocortical projections gave rise to
five distinct cortical fields at locations that preserved the topographic
ordering defined by the original $\gamma$ values. However, we found that
specifying $N$ ordered areas required $M\approx (N+1)/2$ signalling fields. As
the number of distinct guidance fields is unlikely to approach the number of
barrels, further extensions were required.

The term in parentheses in Eq.~\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial density at $t=0$;
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S}  a_i(\mathbf{x}, 0) \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from
Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$. \cmnt{Normalization
  corresponds to the biological constraint that the branching and growth of
  thalamocortical axons is backed by a finite resource provided by their
  parent cell bodies.}

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that a projection interacts with a given ephrin field in
the cortical subplate with a strength determined by the concentration of a
similar molecule at its thalamic origin, i.e., the putative barreloid
center. As such, two orthogonal linear thalamic gradients were defined, from
which 41 pairs of $\gamma$ values were sampled, at the coordinates of 41
barreloid centers recreated from \cite{haidarliu_size_2001} (Fig.\,1B).

A cortical boundary enclosing 41 corresponding barrels was traced from a
cytochrome oxidase stain from \cite{zheng_signal_2001}, and
Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes. From random initial conditions for
$a({\bf x},0)$ and $c({\bf x},0)$, a clear Voronoi-like tesselation emerged
(Fig.\,1C; see Movie S1). A reduction in the Honda delta metric (see
\cite{senft_mouse_1991}) confirmed that a `good' Voronoi pattern emerged
within $\approx 3000$ iterations (see Fig.\,1D). A measure of the difference
in shape between real and simulated barrels revealed a strong correspondence
(see Fig.\,1D).

To further investigate the interplay of genetic and intrinsic mechanisms we
simulated a seminal barrel duplication experiment
\cite{shimogori_fibroblast_2005}, first by solving on a cortical domain
comprising two separate, identically shaped cortical boundaries. In this case
one distinct cortical field per thalamic projection emerged, within one of the
two boundaries only, i.e., no duplication. However, merging the two fields to
create an extended mirror-symmetric boundary shape, with an anterior-posterior
guidance field reflected at the join, gave rise to two mirror-symmetrical
barrel fields comprising 2N barrels, i.e., two cortical fields for each
thalamic projection (Fig.\,1E). Together these results suggest that by the
misexpression of Fgf8, Shimogori \& Grove \citep{shimogori_fibroblast_2005}
effectively created one large barrel field rather than two distinct cortical
fields.

\section{Discussion}

The present results suggest that the key requirements for the emergence of
realistic barrel patterning are i) at each cortical location thalamocortical
projections compete for a limited number of available synaptic connections
(Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a
given projection is reduced by the density of other projections
(Eq.~\ref{eq:comp}), and iii) the branch density of each projection is
conserved over time (Eq.~\ref{eq:norm}).

The emergence of barrels in simulation required competition between thalamic
projections in terms of synaptic connectivity \cmnt{(and an associated
  in-group/out-group axonal identification mechanism)}, and also in terms of
cortical space (as represented by $\chi$). This requirement may account for
the absence of barrels in rodents with larger brains, such as capybara, for
which competition for space is presumably weaker
\citep{woolsey_comparative_1975}. Hence, irrespective of whether barrels are
necessary for adaptive whisker function, the emergence of somatotopically
ordered modular structures may be an inevitable consequence of local
competition for cortical territory driven by input from an array of discrete
sensory organs \citep{purves_iterated_1992}.

\cmnt{The emergence of elongated pre-barrel regions at earlier stages of the
  simulation could be taken as a prediction of the model although where there
  is an absence of the white contours seen in Fig.\,1C/25,000 steps the
  barrels have not yet become well localized and would not be expected to form
  distinct structures.}

It is important to emphasize that the formulation of the model is entirely
local, insofar as simulation requires no information to be communicated from a
given cortical grid cell to any but those immediately adjacent (via
diffusion). Hence the simulations demonstrate how a self-organising system,
constrained by genetically specified guidance cues and by the shape of the
cortical field boundary, can faithfully reproduce an arrangement of cell
aggregates in one neural structure as a topographic map in another.

Moreover, the present results confirm that somatotopic map formation does not
require the pre-specification of centre-points by as yet undetermined
additional developmental mechanisms.

\section{Materials \& Methods}

\cmnt{It's still a short report if I add a condensed version of supp.tex
  here.}

\cmnt{Software for modelling the cortical sheet as a hexagonal lattice with
  no-flux boundary conditions and an arbitrary boundary shape, for computing
  the flux using a finite difference method, for numerical integration by the
  fourth-order Runge-Kutta method, and for reproducing these results is
  available at \url{https://github.com/ABRG-Models/BarrelEmerge}}

\section{Movie S1 caption}

Movie corresponding to Fig.\,1C in the main paper. Simulation parameters were
$N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
$\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
projection for which the connection density is maximal, black lines delineate
boundaries, and overlaid contours show $c>0.5$. The final frame in the movie
is step 25,000 of the simulation.

\section{Acknowledgments}

The authors thank Jason Berwick at the University of Sheffield for advice and
for access to the rat barrel stains used to construct Fig.\,1A. This work was
supported by a Collaborative Activity Award, \emph{Cortical Plasticity Within
  and Across Lifetimes}, from the James S.~McDonnell Foundation (grant
220020516).
